Finiteness in Cubical Type Theory

### 2020-06-18 sections/finiteness-predicates.tex
Update finiteness-predicates.tex

\begin{lemma} \label{split-enum-sigma}
  Split enumerability is closed under \(\Sigma\).
  \begin{equation}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi(x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma(x : A) , U(x))
    }
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a type which is split enumerable, and \(U\) be a type family over
  \(A\) which is split enumerable at every point.
  Formally, we have the following proofs:
  \begin{align}
    \mathcal{E}!_A &: \mathcal{E}!(A) \\
    \mathcal{E}!_U &: \Pi(x : A) , \mathcal{E}!(U(x))
  \end{align}

  Our task is to construct a proof of type:
  \begin{equation}
    \mathcal{E}!(\Sigma(x : A) , U(x))
  \end{equation}
  This proof itself is composed of two components:
  \begin{align}
    \mathit{support} &: \mathbf{List}(\Sigma(x : A) , U(x)) \\
    \mathit{cover} &: \Pi(x : \Sigma(y : A) , U(y)) , x \in \mathit{support}
  \end{align}
  To construct the support list, we apply the function \(\mathcal{E}!_U\) to
  every element in the support list of \(\mathcal{E}!_A\), extract the support
  lists from the resulting finiteness proofs, and concatenate them.

  To prove that this support list does in fact cover the entirety of the type
  \(\Sigma \; A \; U\), we note that any element of type \(\Sigma \; A \; U\)
  must have a first component in the support list of \(\mathcal{E}!_A\), and its
  second component must be in the result of applying \(\mathcal{E}!_U\) to that
  first element (since that support list contains every element of type
  \(U(x)\)).
  Therefore, the pair itself must be in our constructed support list.
\end{proof}

@@

We will now see that split enumerability is in fact a split surjection in
another form:
\begin{lemma} \label{split-enum-is-split-surj}%
  A proof of split enumerability is equivalent to a split surjection from a
  finite prefix of the natural numbers.
  \begin{equation}%
    \mathcal{E}!(A) \simeq \Sigma (n : \mathbb{N}) , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}!(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \in \mathit{xs}
    && \text{def.~\ref{split-enum-def} }(\mathcal{E}!)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{fib}_{\text{snd}(\mathit{xs})}(x)
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{sp-surj-eqn} (sp-surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{sp-surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \twoheadrightarrow ! \; A )
    && \text{eqn.~\ref{sp-surj-arrow-eqn} } (\twoheadrightarrow!) \; \qedhere
  \end{align*}
\end{proof}
\begin{agdalisting}
In our formalisation, the proof is a single line:
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
The only step which isn't definitional equality is the reassociation of
\(\Sigma\).
\ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)

@@

\section{Manifest Bishop Finiteness} \label{manifest-bishop-finiteness}
We mentioned in the introduction that occasionally in constructive mathematics
proofs will contain ``too much'' information.
With split enumerability we can see an instance of this.

\begin{agdalisting} \label{bool-slop}
  Consider the following proof of the finiteness of Bool:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-slop}
  While it represents the ``same'' information as our previous proof, it is
  clearly not the same \emph{object}.
\end{agdalisting}

@@

There is a much simpler and more elegant solution: we insist that every
\emph{membership proof} must be unique.
This would disallow a definition of \(\mathcal{E}!(\AgdaDatatype{Bool})\) with
duplicates, as there are multiple values which inhabit the type \(\text{false}
\in \left[ \text{false}, \text{true}, \text{false} \right]\).
It also allows us to keep most of the split enumerability definition unchanged,
just adding a condition to the returned membership proof in the cover proof.

@@

\begin{lemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{lemma}
This proof takes significantly more work.
The ``unique membership'' condition in \(\mathcal{B}\) means that we are not
permitted duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\mathcal{E}!\) proof: we can do this using the decidable
equality provided by \(\mathcal{E}!\) (lemma~\ref{split-enum-discrete}).
From there, we need to show that the membership proof carries over
appropriately.
\todo{Provide more info on this proof?}

@@

CuTT is not so restricted.
Since we have things like function extensionality and transport, we can indeed
prove the finiteness of function types.
Our proof here makes use directly of the univalence axiom, and makes use
furthermore of all the previous closure proofs.
We will prove this closure on split enumerability, rather than on manifest
Bishop finiteness, as it requires slightly less legwork in the proof itself, but
of course we can derive the proof on manifest Bishop finiteness in a few lines.
\begin{theorem} \label{split-enum-pi}
  Split enumerability is closed under dependent functions.
  (\(\Pi\)-types).
  \begin{equation}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi {(x : A)} , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{theorem}
\begin{proof}
  Let \(A\) be a split enumerable type, and \(U\) be a type family from \(A\),
  which is split enumerable over all points of \(A\).

  As \(A\) is split enumerable, we know that it is also manifestly Bishop finite
  (lemma~\ref{split-enum-to-manifest-bishop}), and consequently we know \(A
  \simeq \mathbf{Fin}(n)\), for some \(n\) (lemma~\ref{bishop-equiv}).
  We can therefore replace all occurrences of \(A\) with \(\mathbf{Fin}(n)\),
  changing our goal to:
  \begin{equation}
    \frac{
      \mathcal{E}!(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\Pi (x : \mathbf{Fin}(n)) , U(x)\right)
    }
  \end{equation}

  We then define the type of \(n\)-tuples over some type family \(T :
  \mathbf{Fin}(n) \rightarrow \mathbf{Type}\).
  \begin{equation}
    \begin{alignedat}{3}
      & \mathbf{Tuple}(0, T)   &&\coloneqq \top \\
      & \mathbf{Tuple}(n+1, T) &&\coloneqq T(0) \times \mathbf{Tuple}(n, T \circ \text{suc})
    \end{alignedat}
  \end{equation}
  We can show that this type is equivalent to functions (proven in our formalisation):
  \begin{equation}
    \Pi(x : \mathbf{Fin}(n)) , U(x) \simeq \mathbf{Tuple}(n, U)
  \end{equation}
  And therefore we can simplify again our goal to the following:
  \begin{equation}
    \frac{
      \mathcal{E}!(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\mathbf{Tuple}(n, U)\right)
    }
  \end{equation}

@@

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\subsection{Closure}
Since we don't have a function of type \(\mathcal{C}(A) \rightarrow
\mathcal{B}(A)\), closure proofs on \(\mathcal{B}\) do not transfer over to
\(\mathcal{C}\) trivially (unlike with \(\mathcal{E}!\) and \(\mathcal{B}\)).
The cases for \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are simple to adapt: we
can just propositionally truncate their Bishop finiteness proof.

@@

  as the one for \(\mathcal{E}!(\top)\).

>>>>

\begin{lemma} \label{split-enum-sigma}
  Split enumerability is closed under \(\Sigma\).
  \begin{equation}
    \frac{
      \AgdaDatatype{\ensuremath{\mathcal{E}!}}(A) \; \; \; \Pi(x : A) , \AgdaDatatype{\ensuremath{\mathcal{E}!}}(U(x))
    }{
      \AgdaDatatype{\ensuremath{\mathcal{E}!}}(\Sigma(x : A) , U(x))
    }
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a type which is split enumerable, and \(U\) be a type family over
  \(A\) which is split enumerable at every point.
  Formally, we have the following proofs:
  \begin{align}
    \AgdaDatatype{\ensuremath{\mathcal{E}!}}_A &: \AgdaDatatype{\ensuremath{\mathcal{E}!}}(A) \\
    \AgdaDatatype{\ensuremath{\mathcal{E}!}}_U &: \Pi(x : A) , \AgdaDatatype{\ensuremath{\mathcal{E}!}}(U(x))
  \end{align}

  Our task is to construct a proof of type:
  \begin{equation}
    \AgdaDatatype{\ensuremath{\mathcal{E}!}}(\Sigma(x : A) , U(x))
  \end{equation}
  This proof itself is composed of two components:
  \begin{align}
    \mathit{support} &: \mathbf{List}(\Sigma(x : A) , U(x)) \\
    \mathit{cover} &: \Pi(x : \Sigma(y : A) , U(y)) , x \in \mathit{support}
  \end{align}
  To construct the support list, we apply the function \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}_U\) to
  every element in the support list of \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}_A\), extract the support
  lists from the resulting finiteness proofs, and concatenate them.

  To prove that this support list does in fact cover the entirety of the type
  \(\Sigma \; A \; U\), we note that any element of type \(\Sigma \; A \; U\)
  must have a first component in the support list of \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}_A\), and its
  second component must be in the result of applying \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}_U\) to that
  first element (since that support list contains every element of type
  \(U(x)\)).
  Therefore, the pair itself must be in our constructed support list.
\end{proof}

@@

We will now see that split enumerability is in fact a split surjection in
another form:
\begin{lemma} \label{split-enum-is-split-surj}%
  A proof of split enumerability is equivalent to a split surjection from a
  finite prefix of the natural numbers.
  \begin{equation}%
    \AgdaDatatype{\ensuremath{\mathcal{E}!}}(A) \simeq \Sigma (n : \mathbb{N}) , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \AgdaDatatype{\ensuremath{\mathcal{E}!}}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \in \mathit{xs}
    && \text{def.~\ref{split-enum-def} }(\AgdaDatatype{\ensuremath{\mathcal{E}!}})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{fib}_{\text{snd}(\mathit{xs})}(x)
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{sp-surj-eqn} (sp-surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{sp-surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \twoheadrightarrow ! \; A )
    && \text{eqn.~\ref{sp-surj-arrow-eqn} } (\twoheadrightarrow!) \; \qedhere
  \end{align*}
\end{proof}
\begin{agdalisting}
In our formalisation, the proof is a single line:
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
The only step which isn't definitional equality is the reassociation of
\(\Sigma\).
\ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)

@@

\section{Manifest Bishop Finiteness} \label{manifest-bishop-finiteness}
We mentioned in the introduction that occasionally in constructive mathematics
proofs will contain ``too much'' information.
With split enumerability we can see an instance of this.
Consider the following proof of the finiteness of Bool:
\begin{agdalisting} \label{bool-slop}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-slop}
\end{agdalisting}
While it represents the ``same'' information as our previous proof, it is
clearly not the same \emph{object}.

@@

There is a much simpler and more elegant solution: we insist that every
\emph{membership proof} must be unique.
This would disallow a definition of \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\;
\AgdaDatatype{Bool}\) with
duplicates, as there are multiple values which inhabit the type \(\text{false}
\in \left[ \text{false}, \text{true}, \text{false} \right]\).
It also allows us to keep most of the split enumerability definition unchanged,
just adding a condition to the returned membership proof in the cover proof.

@@

\begin{lemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{lemma}
This proof takes significantly more work.
The ``unique membership'' condition in \(\mathcal{B}\) means that we are not
permitted duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) proof: we can do this using the decidable
equality provided by \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) (lemma~\ref{split-enum-discrete}).
From there, we need to show that the membership proof carries over
appropriately.
\todo{Provide more info on this proof?}

@@

CuTT is not so restricted.
Since we have things like function extensionality and transport, we can indeed
prove the finiteness of function types.
Our proof here makes use directly of the univalence axiom, and makes use
furthermore of all the previous closure proofs.
We will prove this closure on split enumerability, rather than on manifest
Bishop finiteness, as it requires slightly less legwork in the proof itself, but
of course we can derive the proof on manifest Bishop finiteness in a few lines.
\begin{theorem} \label{split-enum-pi}
  Split enumerability is closed under dependent functions.
  (\(\Pi\)-types).
  \begin{equation}
    \frac{
      \AgdaDatatype{\ensuremath{\mathcal{E}!}}(A) \; \; \; \Pi {(x : A)} , \AgdaDatatype{\ensuremath{\mathcal{E}!}}\left( U(x) \right)
    }{
      \AgdaDatatype{\ensuremath{\mathcal{E}!}}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{theorem}
\begin{proof}
  Let \(A\) be a split enumerable type, and \(U\) be a type family from \(A\),
  which is split enumerable over all points of \(A\).

  As \(A\) is split enumerable, we know that it is also manifestly Bishop finite
  (lemma~\ref{split-enum-to-manifest-bishop}), and consequently we know \(A
  \simeq \mathbf{Fin}(n)\), for some \(n\) (lemma~\ref{bishop-equiv}).
  We can therefore replace all occurrences of \(A\) with \(\mathbf{Fin}(n)\),
  changing our goal to:
  \begin{equation}
    \frac{
      \AgdaDatatype{\ensuremath{\mathcal{E}!}}(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \AgdaDatatype{\ensuremath{\mathcal{E}!}}\left( U(x) \right)
    }{
      \AgdaDatatype{\ensuremath{\mathcal{E}!}}\left(\Pi (x : \mathbf{Fin}(n)) , U(x)\right)
    }
  \end{equation}

  We then define the type of \(n\)-tuples over some type family \(T :
  \mathbf{Fin}(n) \rightarrow \mathbf{Type}\).
  \begin{equation}
    \begin{alignedat}{3}
      & \mathbf{Tuple}(0, T)   &&\coloneqq \top \\
      & \mathbf{Tuple}(n+1, T) &&\coloneqq T(0) \times \mathbf{Tuple}(n, T \circ \text{suc})
    \end{alignedat}
  \end{equation}
  We can show that this type is equivalent to functions (proven in our formalisation):
  \begin{equation}
    \Pi(x : \mathbf{Fin}(n)) , U(x) \simeq \mathbf{Tuple}(n, U)
  \end{equation}
  And therefore we can simplify again our goal to the following:
  \begin{equation}
    \frac{
      \AgdaDatatype{\ensuremath{\mathcal{E}!}}(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \AgdaDatatype{\ensuremath{\mathcal{E}!}}\left( U(x) \right)
    }{
      \AgdaDatatype{\ensuremath{\mathcal{E}!}}\left(\mathbf{Tuple}(n, U)\right)
    }
  \end{equation}

@@

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\subsection{Closure}
Since we don't have a function of type \(\mathcal{C}(A) \rightarrow
\mathcal{B}(A)\), closure proofs on \(\mathcal{B}\) do not transfer over to
\(\mathcal{C}\) trivially (unlike with \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) and \(\mathcal{B}\)).
The cases for \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are simple to adapt: we
can just propositionally truncate their Bishop finiteness proof.

@@

  as the one for \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}(\top)\).
